\huspacy: an industrial-strength Hungarian natural language processing toolkit

### 2022-01-10 mszny2022/main.tex
Camera ready version

\usepackage[utf8]{inputenc}
\usepackage{t1enc}
\usepackage{tikz-dependency}
\usepackage{qtree}
\usepackage{float}
\restylefloat{figure}
\usepackage{tikz-qtree}
\usepackage[english]{babel}
\usepackage{todonotes}
\selectlanguage{english}

\usepackage{array}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{soul}

@@

\renewcommand\todo{\noop}
% \renewcommand\bibname{References}

@@

\author{
György Orosz,
Zsolt Szántó, \break
Péter Berkecz,
Gergő Szabó,
Richárd Farkas\\
\institute{
    Institute of Informatics, University of Szeged\break
    Szeged, Árpád tér 2., H-6720\\
}
\email{gyorgy@orosz.link}\break
\email{\{szantozs,rfarkas\}@inf.u-szeged.hu}\break
}

@@

\begin{abstract}
Although there are a couple of open-source language processing pipelines available for Hungarian, none of them satisfies the requirements of today’s NLP applications. A language processing pipeline should consist of close to state-of-the-art lemmatization, morphosyntactic analysis, entity recognition and word embeddings. 
Industrial text processing applications have to satisfy non-functional software quality requirements, what is more, frameworks supporting multiple languages are more and more favored.
%have been released in large numbers in recent years, such pipelines also have to satisfy non-functional software quality requirements.  
%Last, but not least, the international research and industrial community have been experimenting with languages whose language-specific analyzer tools are available in multilingual language processing libraries. 
This paper introduces \huspacy, an industry-ready Hungarian language processing pipeline. The presented tool provides components for the most important basic linguistic analysis tasks. It is open-source and is available under a permissive license. Our system is built upon spaCy’s NLP components which means that it is fast, has a rich ecosystem of NLP applications and extensions, comes with extensive documentation and a well-known API. Besides the overview of the underlying models, we also present rigorous evaluation on common benchmark datasets. Our experiments confirm that HuSpaCy has high accuracy in all subtasks while maintaining resource-efficient prediction capabilities. 
\end{abstract}

@@

Basic natural language processing tasks such as tokenization, sentence splitting, part-of-speech tagging, lemmatization, dependency parsing, named entity recognition and word embedding representations are amongst the most widely studied problems in natural language processing. Several text analysis applications have been developed during the last decades for both English and other less-resourced languages such as Hungarian. However, a large majority of them solely focus on achieving high accuracy scores on artificial benchmarks and ignore the importance of the underlying model’s practical usability.

@@

Besides the overview of the underlying models, we also  present rigorous evaluation using the Universal grammatical annotations. Our experiments confirm that \huspacy has near state-of-the-art accuracy in many subtasks while maintaining resource-efficient prediction capabilities.

@@

In the early 2010s, Universal PoS \citep{udpos} and Universal Dependency \citep{univedepv1} labeling schemata were developed with the goals of ''cross-linguistically consistent treebank annotation for many languages'' and ''facilitating multilingual parser development, cross-lingual learning, and parsing research from a language typology perspective.'' Many language-specific pipelines changed their representations to the universal annotation schema, but most of them stayed in their own software architecture. Industrial NLP applications are frequently multi-lingual, i.e. the same NLP task has to be solved in several languages. The demand for standardization over languages is high in commercial partners. Beyond universal PoS and dependency annotations, companies who are not NLP experts but want to apply language processing tools prefer multilingual software frameworks and make business decisions to support Hungarian, based on the availability in multilingual frameworks.

The last five years of NLP are dominated by neural language models (NLM) and the applications based on them \citep{nlp-trends}. Academic research has introduced various deep learning methods outperforming previous state of the art in many areas. Such systems usually employ a single neural network providing end-to-end NLP solutions without the need for specific pipeline steps. Also, several pre-trained multilingual NLMs are becoming available which provide standardised solutions for many languages. Regardless of the multilinguality and the high accuracy of deep learning solutions, a lot of critiques have been raised by real-world industrial NLP projects recently.

These are as follows: deep learning solutions often require far more computational resources compared to classic solutions. They heavily rely on GPU acceleration along with significant memory consumption. What is more, their running costs are usually 10 or even 100 times higher than that of alternative solutions. These drawbacks are questioning the commercial return of the accuracy gain. We also note that modern NLP pipelines consist of static word embedding representations and use deep learning for individual pipeline steps as well, hence, the advantage of large neural end-to-end systems
%\footnote{e.g. BERT models} 
are very small.

@@

Considering decades of experience of practical NLP applications, developing an “industrial strength” text processing system is a challenging task. First of all, such a tool should tackle the most important text preprocessing tasks including tokenization, sentence splitting, PoS tagging, lemmatization, dependency parsing, named entity recognition and word embedding representation.

@@

Up until recently, only a few text processing applications were focused on meeting these criteria even for English. When spaCy was released in 2015 \citep{spacy}, it was one one of the first tools targeting mainly real-world industrial applications. Its authors created an unprecedented tool which offers near state-of-the-art accuracy while being an order of magnitude faster than other tools available. SpaCy also comes with an intuitive API, has detailed documentation, and also fits well into the Python ecosystem of machine learning tools. What is more, it is easily deployable and offers built-in syntax and entity visualization tools as well.

@@

There is only one other attempt to provide a unified framework for Hungarian text processing tools: \emtsv{} \citep{emtsv1, emtsv2, emtsv3} (and its predecessor e-magyar \citep{emagyar1, emagyar2}) is a result of a multi-institute collaboration project aiming to integrate existing NLP toolkits into a single application. Unfortunately, neither computational efficiency nor developer ergonomics were amongst the main goals of the project. Although \emtsv{} can yield Universal morphosyntactic annotations through conversion, it is rather inaccurate. What is more, it is not designed to efficiently deal with word vectors, therefore no such facility is available in its system.

@@

\newlength{\lsys}
\settowidth{\lsys}{embeddings}
\begin{table}
\begin{center}
\begin{tabular}{
%lccccc
    l<{\hspace{1em}}
  >{\centering\arraybackslash}m{\lsys}
  >{\centering\arraybackslash}m{\lsys}
  >{\centering\arraybackslash}m{\lsys}
  >{\centering\arraybackslash}m{\lsys}
  >{\centering\arraybackslash}m{\lsys}
}
\toprule
           & NER & 
           \begin{tabular}{@{}c@{}}Word  \\ embeddings\end{tabular} &
           \begin{tabular}{@{}c@{}}High \\ throughput\end{tabular} & 
           \begin{tabular}{@{}c@{}}Part of a \\ multilingual \\ pipeline\end{tabular}
             &
             \begin{tabular}{@{}c@{}}Free for \\ commercial \\ usage\end{tabular} \\
\midrule
\magyarlanc{} & -                       & -                                   & \checkmark                                   & -                                                   & \checkmark                                             \\
\emtsv{}      & \checkmark                       & -                                   & -                                   & -                                                   & -                                             \\
\udpipe{}     & -                       & -                                   & \checkmark                                   & \checkmark                                                   & -                                             \\
\stanza{}     & \checkmark                       & \checkmark                                   & -                                   & \checkmark                                                   & \checkmark                                            
\\
\bottomrule
\end{tabular}
\vspace{1em}
\caption{Hungarian language processing pipelines evaluated with regards of requirements of industrial applicability}
\label{table:systems}
\end{center}
\end{table}

Table \ref{table:systems} summarizes the landscape of the most important Hungarian language processing pipelines and show how they meet the requirements of today’s NLP applications. All in all, we can say that none of them is easily applicable in industrial settings.

We present \huspacy{}, a new industry-ready Hungarian natural language processing toolkit\footnote{We implemented our pipeline in spaCy v3.x. Prior to this work, there were attempts (\url{https://github.com/oroszgy/spacy-hungarian-models}) to support Hungarian in this framework, but none of those releases are compatible with the tool’s current version.}. 
It provides all the aforementioned basic text processing modules with high accuracy. The underlying models are optimized to be light on memory consumption and CPU usage. The presented tool is open source\footnote{\url{https://github.com/huspacy/huspacy}} and is available under the permissive CC-BY-SA-4.0 license. Our system is built on top of spaCy’s infrastructure, thus extensive documentation, debugging tools, an ergonomic API and a flourishing ecosystem are already provided.

\section{Pipeline steps}

This section introduces the NLP algorithms behind \huspacy{}. The presented system is built on spaCy’s architecture, therefore we mainly relied on its symbolic and ML-based text processing infrastructure. The following paragraphs give only a high-level overview of the internals of the parent system and rather concentrate on the contributions of this work.

@@

As regards tokenization, the developed solution uses and extends the underlying system’s algorithm \citep{spacy-tokenization}. It works as follows: first the input text is split on whitespaces, then token boundaries are identified by splitting prefixing or suffixing character sequences. To make this algorithm viable for Hungarian, we extended it with language specific prefix and suffix splitting rules. Furthermore, we had to deal with the ambiguity of tokens around the full stop character, thus a decent abbreviation list has been incorporated to increase the module’s accuracy. During this process we mostly relied on the tests of \huntoken{} \citep{huntoken}.

@@

Sentence boundaries, dependency parse trees, PoS tags, and the corresponding morphosyntactic features are predicted by a multitask deep learning model of the underlying NLP framework. SpaCy’s machine learning approach can be summarized as ''embed, encode, attend, predict''  \citep{spacy-neural-model} which our system adapts for its tagging and parsing components.

Tokens are embedded using the concatenation of static (pretrained) word vectors and ones learned during the task-specific training process. We use a publicly\footnote{\url{https://github.com/oroszgy/hunlp-resources/releases/tag/webcorpuswiki_word2vec_v0.1}} available 300d word embedding which has been trained on the Hungarian Webcorpus \citep{HunWebCorpus} and a snapshot of the Hungarian Wikipedia with CBOW methodology \citep{word2vec}. Task specific word vectors are 256 wide consisting 64 dimensional embeddings of the tokens’ prefixes, suffixes, shapes and the lowercase forms. To make such computation efficient, feature hashing is extensively applied to all kinds of input strings.

During the encoding part, vectors are passed through a four deep stacked CNN encoder \citep{cnn} which uses residual connections and is accompanied with maxout pooling\footnote{The pooling step is considered to be the “attention” mechanism.} \citep{spacy-parser2}. The tagger predicts supertags for tokens which are the combinations of the dependency, part-of-speech and morphosyntactic labels. Efficient prediction is guaranteed by the underlying greedy tagger consisting only of a linear and a softmax layer. As for the dependency parsing, an arc-eager transition system \citep{spacy-parser} is utilized, which shares weights with the tagger model through multitask learning.

Finally, sentence boundary recognition is formalized as a sequence tagging problem where tokens are tagged with a binary label indicating the first token of a sentence. This component is an integral part of the multitask architecture, thus it also shares its lower layers with the parser and the morphosyntactic tagger.

@@

SpaCy’s default lemmatization model is mainly designed for English. It is not suitable for morphologically complex languages such as Hungarian as it only uses lookup tables. Hence, we decided to look for a more sophisticated solution, and integrated the \lemmy{} toolkit \citep{lemmy} which is an open-source Python implementation of the CST rule-learning engine \citep{cst-lemmatizer}. To improve its accuracy we incorporated three minor modifications. First, prefixing numbers of numeric tokens are masked to help the generalization of the engine for inflected numbers. (For example the masked token of `2021-ben’ becomes `0000-ben’.) Second, we enforce lowercase lemmata of sentence starting tokens if they are not proper nouns. Finally, if there are multiple lemma candidates available for a given (word, tag) pair, we pick the one with the highest frequency on the training dataset.

@@

SpaCy’s entity recognizer is built on the transition-based parser architecture described in Section \ref{sec:tagging} (similarly to \citet{spacy-ner}). However, there are two key differences compared to the system of \citet{spacy-ner} The first is that the set of possible transition actions reflects the BILOU tagging scheme. This trick allows the model to have better discrimination ability between different entity classes, furthermore it makes the learning problem easier. Second, the state vector computation includes clues not just from the surrounding words but the tokens of previous entities as well. The sequence tagger model uses BILOU tags for encoding entity boundaries and the decoder is based on a simple greedy softmax layer similar to the morphosyntactic tagger.

\section{Text parsing experiments}

In order to show \huspacy{} abilities, we performed a series of experiments and compared it with the most popular Hungarian NLP pipelines available. For each component, we measured its performance on benchmark datasets along with the model’s throughput and memory consumption. As named entity recognition is usually not part of such pipelines, first we focus on evaluating the text parsing task including tokenization, sentence splitting, morphosyntactic tagging, lemmatization and dependency parsing.

We compare our system with the Hungarian specific \emtsv{} pipeline and two other multilingual toolkits, namely \udpipe{} and \stanza{}. The latter tools score high on parsing UD corpora and both are off-the-shelf available systems. We used these tools as black boxes meaning that no efforts have been made to retrain or fine-tune them. Evaluation is carried out on the test set of the Hungarian subset of the Universal Dependencies Corpus \citep{UniDep} which consists of a small, manually transcribed part of the Szeged Corpus. Performance measures are provided by the evaluation script of the CoNLL 2018 Shared Task\footnote{\url{https://universaldependencies.org/conll18/conll18_ud_eval.py}}.

Up until now, there has  only been a single Hungarian corpus \citep{szegedcorpus} having both morphosyntactic and dependency parse annotations. What is more, UD annotations are available only in a rather small subcorpus of it \citep{UniDep}. While morphosyntactic labels can be transcribed automatically from the Hungarian-specific formalism to UD with high accuracy, the same is not true for dependency parse trees. Hence, there is a notable difference in size of the training data available for tagging and parsing.

To best utilize all possible data sources, we applied a two-step learning strategy
%\footnote{Our experiments are fully reproducible. Hyperparameters of the models lie in configuration files of the given repository.}
using \huspacy\textquotesingle s multitask deep learning model. In the first step, the tagger and the SBD components are pre-trained on the full-size transcribed Szeged Corpus\footnote{When we refer to the Szeged Corpus as a training set, we mean all the sentences that are not part of the development or the test set of the Universal Dependencies corpus.}. This is followed by a fine-tuning step on the gold standard dataset where dependency annotations are also learned. A single step model relying solely on the UD data is also involved in our experiments to allow fair comparison with other tools. 
%%\hl{A summary of hyperparameters can be found in Table X in Appendix X.}

The lemmatizer has been trained with two configurations. First we used only the training set of the Hungarian UD corpus, then we allowed the tool to learn from the whole transcribed Szeged Corpus (except the sentences overlapping with either our test or development sets).

As for benchmarking other pipelines, we only had to evaluate \emtsv, as the authors of \stanza{}\footnote{\url{https://stanfordnlp.github.io/stanza/performance.html}} and \udpipe{}\footnote{\url{https://ufal.mff.cuni.cz/udpipe/1/models}} have already published their tools\textquotesingle \space accuracy on all UD corpora. There are no such published scores available for \emtsv, thus we used the following configuration to produce UD-compatible output for evaluation: \texttt{emToken, emMorph, emLem, emTag, emmorph2ud, emDep, emConll}. While this tool can provide parse trees, their annotation schema is not compatible with that of the Universal Dependencies, hence, its output is not evaluable. We must also note that comparison with the \emtsv{}\textquotesingle s tagger and lemmatizer is not fair, as this tool was trained on a different train-test split which might conflict with ours. (There is a high chance that its training data overlaps with the sentences of our test set.)

\section{Parsing results}

\newlength{\lsz}
\settowidth{\lsz}{Sentence splitting}
\begin{table}
\begin{center}
\begin{tabular}{
  l<{\hspace{1em}}
  >{\centering\arraybackslash}m{\lsz}
  >{\centering\arraybackslash}m{\lsz}
}
\toprule
              & Tokenization             & Sentence splitting \\
\midrule
\stanza{}        & 99.87\%                  & 97.00\%            \\
\udpipe{}        & 99.80\%                  & 95.90\%            \\
\emtsv{}         & 99.77\%                  & 98.67\%            \\
\huspacy{} (UD)  & \multirow{2}{*}{99.89\%} & 99.00\%            \\
\huspacy{} (SZC) &                          & 98.01\%            \\  
\bottomrule
\end{tabular}
\vspace{1em}
\caption{Comparison of the tokenization and sentence boundary detection of NLP pipelines measured in F1 scores on the test of the Hungarian UD Corpus}
\label{table:tokens}
\end{center}
\end{table}

Error propagation is a notable phenomenon in language processing applications, thus using highly accurate preprocessing modules such as tokenization and SBD is indispensable. F1 scores in Table \ref{table:tokens}  suggest that tokenization is easily handled by all of the systems making only a few errors, although \huspacy{} is marginally better compared to the rest of the systems. Sentence boundary detection is a more complex task, where language specific knowledge is necessary. This can be either built into the system (as it is the case with \emtsv{}) or learned by a ML model. Numbers show that both approaches can yield high scoring systems.
%\todo{@Zsolti: honnan lett 99\%-os SBD szamunk?}

>>>>

\usepackage[utf8]{inputenc}
\usepackage{t1enc}
\usepackage{tikz-dependency}
\usepackage{qtree}
\usepackage{float}
\restylefloat{figure}
\usepackage{siunitx}
\usepackage{tikz-qtree}
\usepackage[english]{babel}
\usepackage{todonotes}
% \renewcommand\todo{\noop}
\selectlanguage{english}

\usepackage{array}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{amssymb}
\usepackage{hyperref}
\hypersetup{plainpages=false,
            pageanchor=true,
            colorlinks=true,
            linkcolor=blue,
            citecolor=blue,
            anchorcolor=blue,
            menucolor=blue,
            filecolor=blue,
            urlcolor=blue,
            linktocpage=true,
            breaklinks=true,
            unicode=true}
\usepackage{soul}

@@

% \renewcommand\bibname{References}

@@

\author{
György Orosz,
Zsolt Szántó, \break
Péter Berkecz,
Gergő Szabó,
Richárd Farkas\\
\institute{
    Institute of Informatics, University of Szeged\break
    2. Árpád tér, Szeged, Hungary\\
}
\email{gyorgy@orosz.link}\break
\email{\{szantozs,rfarkas\}@inf.u-szeged.hu}\break
}

@@

\begin{abstract}
Although there are a couple of open-source language processing pipelines available for Hungarian, none of them satisfies the requirements of today’s NLP applications. 
A language processing pipeline should consist of close to state-of-the-art lemmatization, morphosyntactic analysis, entity recognition and word embeddings. 
Industrial text processing applications have to satisfy non-functional software quality requirements, what is more, frameworks supporting multiple languages are more and more favored.
This paper introduces \huspacy, an industry-ready Hungarian language processing toolkit. 
The presented tool provides components for the most important basic linguistic analysis tasks. It is open-source and is available under a permissive license. 
Our system is built upon spaCy’s NLP components resulting in a easily usable, fast yet accurate application.
Experiments confirm that HuSpaCy has high accuracy while maintaining resource-efficient prediction capabilities. 
\end{abstract}

@@

Basic natural language processing tasks such as tokenization, sentence splitting, part-of-speech tagging, lemmatization, dependency parsing and named entity recognition are amongst the most widely studied problems in natural language processing. Several text analysis applications have been developed during the last decades for both English and other less-resourced languages such as Hungarian. However, a large majority of them solely focus on achieving high scores on artificial benchmarks and ignore the importance of practical usability.

@@

First, we give an overview of the underlying models, then rigorous evaluation is presented using various datasets. Finally, experiments are presented confirming that \huspacy{} has high accuracy in many subtasks while maintaining resource-efficient prediction capabilities.

@@

In the early 2010s, Universal PoS \citep{udpos} and Universal Dependency \citep{univedepv1} labeling schemata were developed with the goals of ''cross-linguistically consistent treebank annotation for many languages'' and ''facilitating multilingual parser development, cross-lingual learning, and parsing research from a language typology perspective.'' Many language-specific pipelines changed their representations to these universal annotation schema, but most of them stayed in their own software architecture. Industrial NLP applications are frequently multi-lingual, i.e. the same NLP task has to be solved in several languages. The demand for standardization over languages is high in commercial partners. Beyond universal PoS and dependency annotations, companies who are not NLP experts but want to apply language processing tools prefer multilingual software frameworks and make business decisions to support Hungarian, based on the availability in multilingual frameworks.

The last five years of NLP are dominated by neural language models (NLM) and the applications based on them \citep{nlp-trends}. Academic research has introduced various deep learning methods outperforming previous state of the art in many areas. Such systems usually employ a single neural network providing end-to-end NLP solutions without the need for specific pipeline steps. Also, several pre-trained multilingual NLMs are becoming available which provide standardized solutions for many languages. Regardless of the multilinguality and the high accuracy of deep learning solutions, a lot of critiques have been raised by real-world industrial NLP projects recently.

These are as follows: deep learning solutions often require far more computational resources compared to classic solutions. They heavily rely on GPU acceleration along with significant memory consumption. What is more, their running costs are usually 10 or even 100 times higher than that of alternative solutions. These drawbacks are questioning the commercial return of the accuracy gain. We also note that modern NLP pipelines consist of static word embedding representations and use deep learning for individual pipeline steps as well, hence, the advantage of large neural end-to-end systems might be very small.

@@

Considering decades of experience of practical NLP applications, developing an “industrial-strength” text processing system is a challenging task. First of all, such a tool should tackle the most important text preprocessing tasks including tokenization, sentence splitting, PoS tagging, lemmatization, dependency parsing, named entity recognition and word embedding representation.

@@

Up until recently, only a few text processing applications were focused on meeting these criteria even for English. When spaCy was released in 2015 \citep{spacy}, it was one one of the first tools targeting industrial applications mainly. Its authors created an unprecedented tool which offers near state-of-the-art accuracy while being an order of magnitude faster than other tools available. SpaCy also comes with an intuitive API, has detailed documentation, and also fits well into the Python ecosystem of machine learning tools. What is more, it is easily deployable and offers built-in syntax and entity visualization tools as well.

@@

There is only one other attempt to provide a unified framework for Hungarian text processing tools: \emtsv{} \citep{emtsv1, emtsv2, emtsv3} (and its predecessor e-magyar \citep{emagyar1, emagyar2}) is a result of a multi-institute collaboration project aiming to integrate existing NLP toolkits into a single application. Unfortunately, neither computational efficiency nor developer ergonomics were amongst the main goals of the project. Although \emtsv{} can yield Universal morphosyntactic annotations through conversion, it is rather inaccurate. What is more, it is not designed to efficiently deal with word vectors, therefore no such facility is available in the system.

@@

\newlength{\lsys}
\settowidth{\lsys}{embeddings}
\begin{table}
\begin{center}
\begin{tabular}{
%lccccc
    l<{\hspace{1em}}
  >{\centering\arraybackslash}m{\lsys}
  >{\centering\arraybackslash}m{\lsys}
  >{\centering\arraybackslash}m{\lsys}
  >{\centering\arraybackslash}m{\lsys}
  >{\centering\arraybackslash}m{\lsys}
}
\toprule
           & NER & 
           \begin{tabular}{@{}c@{}}Word  \\ embeddings\end{tabular} &
           \begin{tabular}{@{}c@{}}High \\ throughput\end{tabular} & 
           \begin{tabular}{@{}c@{}}Part of a \\ multilingual \\ pipeline\end{tabular}
             &
             \begin{tabular}{@{}c@{}}Free for \\ commercial \\ usage\end{tabular} \\
\midrule
\magyarlanc{} & -                       & -                                   & \checkmark                                   & -                                                   & \checkmark                                             \\
\emtsv{}      & \checkmark                       & -                                   & -                                   & -                                                   & -                                             \\
\udpipe{}     & -                       & -                                   & \checkmark                                   & \checkmark                                                   & -                                             \\
\stanza{}     & \checkmark                       & \checkmark                                   & -                                   & \checkmark                                                   & \checkmark                                            
\\
\bottomrule
\end{tabular}
\vspace{1em}
\caption{Hungarian language processing pipelines evaluated with regards of requirements of industrial applicability}
\label{table:systems}
\end{center}
\vspace{-2em}
\end{table}

Table \ref{table:systems} summarizes the landscape of the most important Hungarian language processing pipelines and show how they meet the requirements of today’s NLP applications. We must note that \emtsv{} does not have any restriction for using it in commercial applications, although some of its most important components have very restrictive licenses (e.g. \texttt{emMorph}). All in all, we can say that none of them is easily applicable in industrial settings.

We present \huspacy{}, a new industry-ready Hungarian natural language processing toolkit.
%\footnote{We implemented our pipeline in spaCy v3.x. Prior to this work, there were attempts (\url{https://github.com/oroszgy/spacy-hungarian-models}) to support Hungarian in this framework, but none of those releases are compatible with the tool’s current version.}. 
It provides all the aforementioned basic text processing modules with high accuracy. The underlying models are optimized to be light on memory consumption and CPU usage. The presented tool is open source\footnote{\url{https://github.com/huspacy/huspacy}} and is available under the permissive CC-BY-SA-4.0 license. Our system is built on top of spaCy’s infrastructure, thus extensive documentation, debugging tools, an ergonomic API and a flourishing ecosystem are already provided.

\section{HuSpaCy internals}

This section introduces the NLP algorithms behind the presented tool. As our system is built on spaCy’s architecture, we mainly relied on its symbolic and ML-based text processing infrastructure. The following paragraphs give a high-level overview of the framework utilized and also describes the contributions of this work.

@@

\huspacy{} builds on spaCy\textquotesingle s \citep{spacy-tokenization} tokenization infrastructure which works as follows: first the input text is split on whitespaces, then token boundaries are identified by splitting prefixing or suffixing character sequences. To make this algorithm viable for Hungarian, we extended it with language specific prefix and suffix splitting rules. Furthermore, we had to deal with the ambiguity of tokens around full stops, thus an extensive abbreviation list has been incorporated to increase the module\textquotesingle s accuracy. During this process we mostly relied on the test cases of \huntoken{} \citep{huntoken} to fine-tune the algorithm.

@@

Sentence boundaries, dependency parse trees, PoS tags, and the corresponding morphosyntactic features are predicted by a multitask deep learning model of the underlying NLP framework.  SpaCy\textquotesingle s machine learning approach can be summarized as ''embed, encode, attend, predict''  \citep{spacy-neural-model} which our system adapts for its tagging and parsing components.

Tokens are embedded using the concatenation of static (pretrained) word vectors and ones learned during the task-specific training process. 
We use a publicly\footnote{\url{https://github.com/oroszgy/hunlp-resources/releases/tag/webcorpuswiki_word2vec_v0.1}} available 300d word embedding which has been trained on the Hungarian Webcorpus \citep{HunWebCorpus} and a snapshot of the Hungarian Wikipedia with CBOW methodology \citep{word2vec}. 
Task specific word vectors are 256 wide consisting 64 dimensional embeddings of the tokens\textquotesingle{} prefixes, suffixes, shapes and the lowercase forms. To make such computation efficient, feature hashing is extensively applied to all kinds of input strings.

During the encoding part, vectors are passed through a four deep stacked CNN encoder \citep{cnn} which uses residual connections and is accompanied with maxout pooling\footnote{The pooling step is considered to be the “attention” mechanism.} \citep{spacy-parser2}. 
%The tagger predicts supertags for tokens which are the combinations of the dependency, part-of-speech and morphosyntactic labels. 
Efficient prediction is guaranteed by the underlying greedy tagger consisting only of a linear and a softmax layer. As for the dependency parsing, an arc-eager transition system \citep{spacy-parser} is utilized, which shares weights with the tagger model through multitask learning.

Finally, sentence boundary recognition is formalized as a sequence tagging problem where tokens are tagged with a binary label indicating the first token of a sentence. This component is an integral part of the multitask architecture, thus it also shares its neural model with the parser and the morphosyntactic tagger.

@@

SpaCy\textquotesingle s default lemmatization model is mainly designed for English. It is not suitable for morphologically complex languages such as Hungarian as it only uses lookup tables. Hence, we decided to look for a more sophisticated solution and adapted the \lemmy{} toolkit \citep{lemmy} which is an open-source Python implementation of the CST rule-learning engine \citep{cst-lemmatizer}. To improve its accuracy we incorporated three minor modifications. First, prefixing numbers of numeric tokens are masked to help the engine in case of inflected numbers. (For example the masked token of `2021-ben’ becomes `0000-ben’.) Second, we enforce lowercasing of sentence starting tokens if they are not proper nouns. Finally, if there are multiple lemma candidates available for a given (word, tag) pair, we pick the one with the highest frequency on the training dataset.

@@

SpaCy’s entity recognizer is built on the transition-based parser architecture described in Section \ref{sec:tagging} (similarly to \citet{spacy-ner}). 
However, there are two key differences compared to the system of \citet{spacy-ner}. 
The first is that the set of possible transition actions reflects the BILOU tagging scheme. This trick allows the model to have better discrimination ability between different entity classes, furthermore it makes the learning problem easier. 
Second, the state vector computation includes clues not just from the surrounding words but the tokens of previous entities as well. 
The sequence tagger model uses BILOU tags for encoding entity boundaries and the decoder is built on a greedy softmax layer similar to that of the morphosyntactic tagger.

\section{Experiments and results}
\subsection{Text parsing}
\label{parser-eval}

In order to benchmark \huspacy{}, we performed a series of experiments comparing its performance with the most popular off-the-shelf pipelines available. Evaluation is carried out on the test set of the Hungarian Universal Dependencies Corpus \citep{UniDep} by using the evaluation script of the CoNLL 2018 Shared Task\footnote{\url{https://universaldependencies.org/conll18/conll18_ud_eval.py}}.

%It is worth noting that this dataset is a manually transcribed part of the Szeged Corpus.

% / F1-score of each text procesin performance on benchmark datasets along with the model’s throughput and memory consumption. As named entity recognition is usually not part of such pipelines, first we focus on evaluating the text parsing task including tokenization, sentence splitting, morphosyntactic tagging, lemmatization and dependency parsing.

Three popular text processing tools have been selected for comparison. \emtsv{} is a Hungarian specific pipeline integrating state-of-the-art NLP components, \udpipe{} is used as a baseline system in CoNLL competitions, while \stanza{} has high scores on parsing UD corpora. All systems are used  as black boxes meaning they have not been retrained or fine-tuned.

% and two other multilingual toolkits, namely \udpipe{} and \stanza{}. The latter tools score high on parsing UD corpora and both are off-the-shelf available systems. We used these tools as black boxes meaning that no efforts have been made to retrain or fine-tune them. Evaluation is carried out on the test set of the Hungarian subset of the Universal Dependencies Corpus \citep{UniDep} which consists of a small, manually transcribed part of the Szeged Corpus. Performance measures are provided by the evaluation script of the CoNLL 2018 Shared Task\footnote{\url{https://universaldependencies.org/conll18/conll18_ud_eval.py}}.

Up until now, there has only been a single Hungarian corpus \citep{szegedcorpus} having both morphosyntactic and dependency parse annotations. What is more, UD annotations are available only in a rather small subcorpus of it \citep{UniDep}. As PoS and morphosyntactic labels can be transcribed automatically from the Hungarian-specific formalism to UD with high accuracy, additional silver standard data can be utilized to train taggers.

In case of \huspacy{}, we applied a two-step learning strategy
\footnote{Hyperparameters of the models are available in the tool\textquotesingle s repository (tag \texttt{v0.4.2}) as configuration files.} to best utilize all available training data. 
In the first step, the tagger and the SBD components are pre-trained on the whole transcribed SZC\footnote{When we refer to the Szeged Corpus as a training set, we mean all the sentences that are not part of the development or the test set of the Universal Dependencies corpus.}. This is followed by a fine-tuning step on the gold standard UD dataset where dependency annotations are also learned. 
To allow fair comparison with \stanza{} and \udpipe{}, a single step model relying solely on the UD data is also involved in the evaluation. 
%%\hl{A summary of hyperparameters can be found in Table X in Appendix X.}

For similar reasons, the lemmatizer has been trained with two configurations. First we used only the training set of the Hungarian UD corpus, then we allowed the tool to learn from the whole transcribed Szeged Corpus (except the sentences overlapping with either our test or development sets).

The authors of \stanza{}\footnote{\url{https://stanfordnlp.github.io/stanza/performance.html}} and \udpipe{}\footnote{\url{https://ufal.mff.cuni.cz/udpipe/1/models}} have already published their tools\textquotesingle \space accuracy on the Hungarian UD corpus, however the same is not true for \emtsv{.}
To evaluate the latter toolkit we used the following (default) configuration to produce an UD-compatible output: \texttt{emToken, emMorph, emLem, emTag, emmorph2ud, emDep, emConll}. 
While \emtsv{} can provide parse trees, their annotation schema is not compatible with that of the Universal Dependencies, hence, its output is not evaluable. 
We must also note that comparison with the \emtsv{}\textquotesingle s tagger and lemmatizer might not be fair, as this tool was trained on a different train-test split which might conflict with ours. (There is a high chance that its training data overlaps with the sentences of our test set.)

\newlength{\lsz}
\settowidth{\lsz}{Sentence splitting}
\begin{table}
\begin{center}
\begin{tabular}{
  l<{\hspace{1em}}
  >{\centering\arraybackslash}m{\lsz}
  >{\centering\arraybackslash}m{\lsz}
}
\toprule
              & Tokenization             & Sentence splitting \\
\midrule
\stanza{}        & 99.87\%                  & 97.00\%            \\
\udpipe{}        & 99.80\%                  & 95.90\%            \\
\emtsv{}         & 99.77\%                  & 98.67\%            \\
\huspacy{} (UD)  & \multirow{2}{*}{99.89\%}     & 97.66\%            \\
\huspacy{} (SZC) &                          & 97.54\%            \\  
\bottomrule
\end{tabular}
\vspace{1em}
\caption{Tokenization and sentence boundary detection F1 scores on the test of the Hungarian UD Corpus}
\label{table:tokens}
\end{center}
\vspace{-3em}
\end{table}

% Error propagation is a notable phenomenon in language processing applications, thus using highly accurate preprocessing modules such as tokenization and SBD is indispensable. 
F1 scores in Table \ref{table:tokens}  suggest that tokenization is easily handled by all of the systems, although \huspacy{} is marginally better compared to the rest of the tools. 
Sentence boundary detection is a more complex task, where language specific knowledge is necessary. This can be either built into the system (as it is the case with \emtsv{}) or learned by a ML model. Numbers show that both approaches can yield satisfactory SBD components, although the rule-based solution of \emtsv{} stands out followed by the tagging approach of our pipeline.

\newlength{\lm}
\settowidth{\lm}{Morph. acc.}
\begin{table}
\begin{center}
\begin{tabular}{
    l<{\hspace{1em}}
  >{\centering\arraybackslash}m{\lm}
  >{\centering\arraybackslash}m{\lm}
  >{\centering\arraybackslash}m{\lm}
  >{\centering\arraybackslash}m{\lm}
}
\toprule
              & PoS acc. & Morph. acc. & UAS   & LAS   \\
\midrule
\stanza{}        & 96.03\%      & 93.76\%               & 83.62\%   & 78.86\%   \\
\udpipe{} v1     & 90.60\%      & 88.50\%               & 72.80\%   & 67.20\%   \\
\emtsv{}         & 89.19\%      & 89.12\%               & \multicolumn{2}{c}{--} \\
\huspacy{} (UD)  & 94.70\%      & 89.03\%               & 79.03\%   & 73.17\%   \\
\huspacy{} (SZC) & 96.58\%      & 93.23\%               & 79.39\%   & 74.22\%   \\
\bottomrule
\end{tabular}
\vspace{1em}
\caption{Comparison of  tagging accuracy and attachment scores of the benchmarked pipelines on the test set of the Hungarian UD Corpus.}
\label{table:tagging}
\end{center}
\vspace{-3em}
\end{table}

Tagging accuracy and attachment scores are presented in Table \ref{table:tagging}. 
Results show that \stanza{} is a clear winner in dependency parsing while the PoS tagging score of \huspacy{} (the one using additional training data) is the highest one. 
It can be seen that the usage of the extra silver standard data yields better performance for our models both during tagging and dependency parsing. 
\udpipe{} and \emtsv{} have relatively low scores: 
the results of \udpipe{} are not surprising (cf. \cite{conll-2018}), but  \emtsv{}\textquotesingle s scores are unexpected given that it is built upon state-of-the-art morphosyntactic tagging facilities \citep{purepos}. 
%Our assumption is that there could be tag conversion compatibility issues which could cause such low scores. However, this hypothesis needs further investigation.
